Empirical Bayes analysis of \RNAseq{} data for detection of gene expression heterosis

### 2015-09-01 response.tex

\response{Added at the end of the Discussion section.}

\comment{I realize that Ji et al.’s (2014) work was intended for microarray data, or more specifically Gaussian likelihoods, but there has been some work that suggests that Gaussian likelihood specifications in MAANOVA may, oddly enough, have better properties for the analysis of RNA-seq data than more popular methods specifically intended for count data (edgeR and DESeq).  See Reeb and Steibel (2013) and also comments from one reviewer along these lines.  Since Ji et al.’s work also derives from ISU, have you assessed its performance relative to your proposed method on RNA-seq data?}

\response{We have included Ji et al.'s work as another alternative method, see Sections 3.2 and 3.3.}

\comment{One seemingly glaring omission from this work is any comparative assessment of the proposed versus other methods (e.g. ShrinkBayes) on the actual maize dataset in Section 4.  Why is that missing? The comments made by one reviewer regarding a relative assessment of computing time relative to the other methods deserves attention.}

\response{We have provided a comparison of the computational time on the real data set using these methods. We have compared point estimates from edgeR to those using eBayes (laplace). }

@@

\response{The area under the curve attempts to do exactly this, i.e. a single metric to compare the methods. We do not try to over-emphasize the difference between the ShrinkBayes and eBayes methods, which are both based on our model, as we do not believe there is much of a difference.}

Reviewer \#4

\comment{1. The authors propose a method to assess "gene expression heterosis."   While it is mentioned that this type of heterosis is a possible explanation for phenotypic heterosis, I'd like more discussion regarding this point.  The two references seem a bit outdated (for bioinformatics) so I'm wondering if this explanation is still being considered.  If it is, than I'd like to see more discussion about how the analytic results would be used to investigate this explanation.   Wouldn't scientists have specific phenotypic traits in mind and know of genes associated with these traits?  I'm a little confused why the search through the whole genome for this type of heterosis if they're looking to draw inference on the association.}

\response{We now include a reference to Chen (2013), a Nature Reviews Genetics article entitled ``Genomic and epigenetic insights into the molecular bases of heterosis''. Even though scientists may have specific candidates. An exploratory analysis to suggest other genes seems useful.}

### 2015-09-01 ebayes_rnaseq_heterosis.tex
computation time

We analyzed the data using all the methods compared in the previous section. The computation time on a Dell Precision T3600 with a 3.6GHz intel Xeon processor was 14 seconds for \edgeR{}, 1.3 minutes for \citeauthor{ji2014estimation}, 10 hours for the eBayes approaches, and 17 hours for \baySeq{}. In the eBayes approach, the vast majority of the time is spent on independent MCMC analysis for each gene. Thus we parallelized this step using {\tt doMC} \citep{revolution2014domc} and {\tt plyr} \citep{hadley2011split}. When parallelized across 5 cores, the eBayes approaches took about 2.5 hours.

@@

Figure \ref{f:gene_specific_estimates} provides point estimates of the gene-specific parameters from the initial \edgeR{} estimation step and after the eBayes with Laplace distribution procedure described in Section \ref{s:ebayes}. 
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{gene_specific_estimates.pdf}}
\caption{Two-dimensional histogram of point estimates from \edgeR{} and posterior means from eBayes using Laplace distributions along with the $y=x$ diagonal.}
\label{f:gene_specific_estimates}
\end{figure}
This figure shows shrinkage for large absolute values of $\alpha$ and $\delta$ from the \edgeR{} estimation toward $\hat{\eta}_\alpha$ and $\hat{\eta}_\delta$ which are both approximately zero. The figure also shows decreased values for the overdispersion parameter with larger decreases for high and low values of overdispersion. Finally, very little, if any, shrinkage is observed for $\phi$.

@@

Figure \ref{f:volcano} provides a volcano plot, in this case a bivariate histogram, to visualize the maximum of the probabilities of LPH and HPH versus estimated effect size. 
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{volcano}}
\caption{A bivariate histogram of the maximum of the LPH and HPH probabilities versus estimated effect size defined in equation \eqref{e:effect_size} for the B73 $\times$ Mo17 maize experiment.}
\label{f:volcano}
\end{figure}
The figure shows a ridge at an effect size of zero for probabilities below 0.5. Above a probability of 0.5, we see a prototypical volcano pattern, with increased EH probability corresponding to larger estimated effect sizes and no estimated effect sizes near zero for genes with high EH probability. We also see asymmetry, with larger negative effect sizes than positive effect sizes due to genes with hybrid counts near zero and relatively high parental counts. Genes with high estimated posterior probabilities of EH and large estimated effect sizes are candidates for further investigation.

>>>>

We analyzed the data using all the methods compared in the previous section. The computation time on a Dell Precision T3600 with a 3.6GHz Intel Xeon processor was 14 seconds for \edgeR{}, 1.3 minutes for \citeauthor{ji2014estimation}, 10 hours for the eBayes approaches, and 17 hours for \baySeq{}. In the eBayes approach, the vast majority of the time is spent on independent MCMC analysis for each gene. Thus we parallelized this step using {\tt doMC} \citep{revolution2014domc} and {\tt plyr} \citep{hadley2011split}. When parallelized across 5 cores, the eBayes approaches took about 2.5 hours.

@@

Figure \ref{f:gene_specific_estimates} provides point estimates of the gene-specific parameters from the initial \edgeR{} estimation step and after the eBayes with Laplace distribution procedure described in Section \ref{s:ebayes}. 
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{gene_specific_estimates.pdf}}
\caption{Two-dimensional histogram of point estimates from \edgeR{} and posterior means from eBayes (Laplace) along with the $y=x$ diagonal.}
\label{f:gene_specific_estimates}
\end{figure}
This figure shows shrinkage for large absolute values of $\alpha$ and $\delta$ from the \edgeR{} estimation toward $\hat{\eta}_\alpha$ and $\hat{\eta}_\delta$ which are both approximately zero. The figure also shows decreased values for the overdispersion parameter with larger decreases for high and low values of overdispersion. Finally, very little, if any, shrinkage is observed for $\phi$.

@@

Figure \ref{f:volcano} provides a volcano plot, in this case a bivariate histogram, to visualize the maximum of the probabilities of LPH and HPH versus estimated effect size. 
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{volcano}}
\caption{A bivariate histogram of the maximum of the LPH and HPH probabilities versus estimated effect size defined in equation \eqref{e:effect_size} for the B73 $\times$ Mo17 maize experiment using eBayes (Laplace).}
\label{f:volcano}
\end{figure}
The figure shows a ridge at an effect size of zero for probabilities below 0.5. Above a probability of 0.5, we see a prototypical volcano pattern, with increased EH probability corresponding to larger estimated effect sizes and no estimated effect sizes near zero for genes with high EH probability. We also see asymmetry, with larger negative effect sizes than positive effect sizes due to genes with hybrid counts near zero and relatively high parental counts. Genes with high estimated posterior probabilities of EH and large estimated effect sizes are candidates for further investigation.
